---
title: "Get started with batchit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with batchit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem batchit solves

A long loop in one R session can accumulate memory that it never gives back. R
may keep a large allocation after the objects that used it are gone. The loop
then carries that memory to the end of the run. Letting a process exit is the
reliable way to reclaim it.

batchit runs one R function once per item, and gives each call its own worker
process. The worker process exits when its item finishes, and its memory goes
back to the operating system. This matters when one item peaks at many
gigabytes.

Four functions do that dispatch. They share one shape. They differ on two axes:
how each call's arguments reach the worker process, and what comes back.

Most examples in this article run when R builds the article, and show their real
output. Two kinds do not run. The streaming example needs a package that you
install yourself. The self-containedness fragments are there to be read.

## Your first dispatch

Square three numbers, in separate R processes, two at a time.

```{r first}
library(batchit)

results <- run_and_collect(
  fn = function(x) list(value = x^2, worker_pid = Sys.getpid()),
  items = list(list(x = 2), list(x = 3), list(x = 4)),
  n_workers = 2
)

results[[1]]
```

batchit prints one dispatch line, and then one progress line per completed item.
Most later examples in this article hide those lines, to keep the output short.

Each item also reported the process id it ran in. Compare those three ids with
this session's id:

```{r first-pids, message = FALSE}
worker_pids <- vapply(results, function(r) r$worker_pid, integer(1))
worker_pids

Sys.getpid()
Sys.getpid() %in% worker_pids
```

batchit started a fresh R process for each of the three items, and none of them
was this session.

## What just happened

Five terms carry through every batchit call.

**`fn`** is the function to run once per item.

**An item is one call's arguments, as a named list.** `list(x = 2)` means "call
`fn` with `x = 2`". Three of the four dispatch functions take a list of items, as
`items`.

**`n_workers`** is how many items run at the same time.

**A worker process is a separate R process.** `run()`, `run_and_collect()` and
`run_and_write_files_atomically()` start a fresh one for every item.
`stream_from_parent_and_write_files_atomically()` instead reuses a small pool of
persistent worker processes (`mirai` daemons).

**The parent R session** is the session you call batchit from. This article uses
that term throughout.

## Which of the four do I want?

Two questions decide it.

**What do you want back?** Pick one of three answers:

- each item's return value;
- nothing, because `fn` has a side effect outside R;
- output files that batchit writes and guards for you.

**Can you hold every item in memory at once?** Usually yes. Say no when one
item's arguments are a large data slice. Say no when there are far too many items
to keep as one list.

| What you want back | Items | Use |
|---|---|---|
| each item's return value | all in memory | `run_and_collect()` |
| nothing | all in memory | `run()` |
| output files batchit writes | all in memory | `run_and_write_files_atomically()` |
| output files batchit writes | produced one at a time | `stream_from_parent_and_write_files_atomically()` |

Lazy production is available only in combination with batchit-managed output
files. There is no streaming function that hands back raw return values.

Consider the simpler alternative before you reach for the streaming function. Put
a small identifier or file path in each item. Have `fn` load the large data
itself, inside its worker process. That keeps the fresh-process-per-item memory
behaviour, which the streaming function gives up.

## `run_and_collect()`: a parallel `lapply()`

Use this when you need every item's return value.

```{r collect, message = FALSE}
squares <- run_and_collect(
  fn = function(x) x^2,
  items = list(small = list(x = 2), medium = list(x = 3), large = list(x = 4)),
  n_workers = 2
)

squares
```

The returned list carries no names, even though `items` was named `small`,
`medium` and `large` here. The two file-writing functions below do name their
result, so this difference is worth remembering.

Results also come back **in the order of `items`**, not in the order the worker
processes finished. The first item below sleeps for a second and still lands
first:

```{r collect-order, message = FALSE}
run_and_collect(
  fn = function(label, seconds) {
    Sys.sleep(seconds)
    label
  },
  items = list(
    list(label = "slow", seconds = 1),
    list(label = "fast", seconds = 0)
  ),
  n_workers = 2
)
```

Every return value travels from the worker process back to the parent R session.
A small object can also travel into a worker process as part of an item. For a
large object, it is usually better to have `fn` read it from disk itself.

## `run()`: run items and discard the return values

Use this when `fn` is called for an effect outside R. It writes its own files, or
it updates a database. It works like `run_and_collect()` and returns
`invisible(NULL)`.

The effect has to be external. Each item runs in a separate R process. An
assignment to a global variable inside `fn` changes that worker process, not the
parent R session.

```{r run, message = FALSE}
run_dir <- file.path(tempdir(), "batchit-run")
dir.create(run_dir, showWarnings = FALSE)

run(
  fn = function(x, dir) qs2::qs_save(x^2, file.path(dir, paste0(x, ".qs2"))),
  items = list(
    list(x = 1, dir = run_dir),
    list(x = 2, dir = run_dir),
    list(x = 3, dir = run_dir)
  ),
  n_workers = 2
)

list.files(run_dir)
qs2::qs_read(file.path(run_dir, "2.qs2"))
```

Files that `fn` writes on its own, as here, get no protection from batchit. An
interruption partway through writing `2.qs2` leaves a half-written `2.qs2` at
that path. `run_and_write_files_atomically()` is what prevents that.

## `run_and_write_files_atomically()`: batchit writes the files

Declare, per item, the final path of every file that item must produce. batchit
writes each one to a temporary file, in the same directory as its destination. It
replaces the final files only after every one of that item's outputs is written
successfully.

`outputs` is one named character vector per item, in the same order as `items`.
When `items` is named, `outputs` MAY carry the same names instead, in any order.

Three rules apply to every declared path:

- the path MUST be absolute;
- its parent directory MUST already exist;
- the path MUST be unique across the whole call.

A destination MAY already exist as a plain file, and batchit then replaces it.
batchit rejects a directory or a symlink at a destination. Base R cannot portably
reject a FIFO, socket or device file, so batchit does not catch those.

### `style = "return"` (the default)

`fn` returns a named list, one element per declared output. Its names MUST be the
same set as the declared output names, in any order. A duplicate name, a missing
name and an extra name are all errors. batchit saves each value to its path in
the `qs2` format.

```{r write-return, message = FALSE}
write_dir <- file.path(tempdir(), "batchit-write")
dir.create(write_dir, showWarnings = FALSE)

record <- run_and_write_files_atomically(
  fn = function(x) list(squared = x^2, doubled = x * 2),
  items = list(two = list(x = 2), three = list(x = 3)),
  outputs = list(
    two = c(
      squared = file.path(write_dir, "sq_two.qs2"),
      doubled = file.path(write_dir, "db_two.qs2")
    ),
    three = c(
      squared = file.path(write_dir, "sq_three.qs2"),
      doubled = file.path(write_dir, "db_three.qs2")
    )
  ),
  n_workers = 2
)

qs2::qs_read(file.path(write_dir, "sq_three.qs2"))
```

The return value is not `fn`'s return value. It is a record of what batchit
wrote. The record holds one element per item, in the order of `items`, named by
item id:

```{r write-record}
str(record)
```

`committed` maps each declared output name to the final path batchit wrote.
`attempt` is an internal per-item identifier; ignore it.

### `style = "staged_writer"`: `fn` writes the files itself

Ask `where_to_write_output(name)` for the path to write each declared output to,
and write there. batchit discards `fn`'s return value without examining it.

```{r write-staged, message = FALSE}
staged <- run_and_write_files_atomically(
  fn = function(x) {
    qs2::qs_save(x^2, batchit::where_to_write_output("squared"))
    writeLines(as.character(x * 2), batchit::where_to_write_output("doubled"))
    invisible(NULL)
  },
  items = list(two = list(x = 2), three = list(x = 3)),
  outputs = list(
    two = c(
      squared = file.path(write_dir, "sq2.qs2"),
      doubled = file.path(write_dir, "db2.txt")
    ),
    three = c(
      squared = file.path(write_dir, "sq3.qs2"),
      doubled = file.path(write_dir, "db3.txt")
    )
  ),
  style = "staged_writer",
  n_workers = 2
)

readLines(file.path(write_dir, "db2.txt"))
```

A discarded return value does not mean batchit takes your word for it. Before it
commits, batchit checks that every declared output now exists at its staging
path, as a plain, non-symlink file. It errors and names the output that is
missing.

Two rules go with `where_to_write_output()`:

1. Write to the path it returns, and otherwise leave that path alone. batchit
   moves the file into place itself.
2. Call it as `batchit::where_to_write_output()` from an inline function. An
   inline function MAY only call another package in package-qualified form.

Pick `"staged_writer"` when `fn` already writes files on its own. That covers a
format other than `qs2`, and another package's writer. `fn` then writes each file
directly, instead of building the whole object in memory for batchit to
serialize. Pick `"return"` the rest of the time, which is most of the time.

## `stream_from_parent_and_write_files_atomically()`: build items lazily

Use this when building the full `items` list would itself use too much memory.
Instead of `items`, give an `ids` vector and a `producer(id)` function that builds
one item.

batchit keeps at most `min(2 * n_workers, length(ids))` items in flight, and calls
`producer()` only when one of those slots is free. A slot is not the same as an
idle worker process. When every worker process is busy, roughly `n_workers`
further items can already be produced and queued.

The bound covers **produced item arguments only**. The `ids` vector, the
`outputs` list and the accumulating result list live in the parent R session for
the whole call. `n_workers` bounds none of those three. Streaming solves the
problem of items too large to hold all at once. It does not make the per-item
bookkeeping free.

This example does not run here, because it needs a function inside a package that
you install yourself.

```{r stream, eval = FALSE}
# `write_one_slice()` must live in an INSTALLED package: this function loads it
# by package name and function name, never by value. Put it in your own
# package's R/ directory, install, and replace "yourpkg" below.
#
#   write_one_slice <- function(slice) list(main = slice)

stream_dir <- file.path(tempdir(), "batchit-stream")
dir.create(stream_dir, showWarnings = FALSE)
ids <- c("a", "b", "c")

stream_from_parent_and_write_files_atomically(
  fn = package_function("yourpkg", "write_one_slice"),
  ids = ids,
  # Replace this with your own loader; it runs in the parent R session.
  producer = function(id) list(slice = toupper(id)),
  outputs = stats::setNames(
    lapply(ids, function(id) c(main = file.path(stream_dir, paste0(id, ".qs2")))),
    ids
  ),
  n_workers = 2
)
```

`producer()` runs in the parent R session, never in a worker process, so it is an
ordinary function. The self-containedness rules below apply to `fn`, not to it.
Load or build each item's data inside `producer()`. Doing it beforehand defeats
the point.

Once `producer()` returns an item, batchit applies the same `outputs` validation,
the same two `style`s and the same commit record as
`run_and_write_files_atomically()`.

Three differences from the other three functions.

1. `fn` MUST come from `package_function()`. An inline function is not accepted.
2. It needs the `mirai` package: `pak::pak("mirai")`.
3. Its worker processes are persistent `mirai` daemons. Their memory is **not**
   reclaimed between items, the way process exit reclaims it elsewhere. So this
   function suits many small-to-medium items better than a few huge ones.

Those daemons live in a private compute profile that batchit creates and tears
down for the one call. It never touches a `mirai` daemon configuration you set up
yourself.

## Naming the function: inline or `package_function()`

An **inline function** is an ordinary R function, passed by value. Write it
directly in the call, or assign it to a variable first. It is the quick option,
and every example above uses it, except the streaming one.

It must be self-contained. It may use only its own arguments, base R, and
`pkg::fun()`-qualified calls to other packages. The four fragments below are
there to be read, so they do not run:

```{r inline-rules, eval = FALSE}
# Allowed, own argument and base R:
function(x) x^2

# Allowed, another package, package-qualified:
function(path) qs2::qs_read(path)

# NOT allowed, `threshold` is not an argument of this function:
threshold <- 10
function(x) x > threshold

# NOT allowed, `my_helper` is a plain call to a function defined outside:
my_helper <- function(x) x * 2
function(x) my_helper(x)
```

batchit checks this with a best-effort static scan, not a proof. An indirect
lookup can pass the check and still fail in the worker process. `get()`,
`eval(parse())` and a string-based `do.call()` are all indirect lookups. They
fail because batchit replaces the function's environment before it sends the
function. Keep to the rule even where the check lets something through.

If `fn` needs something from outside its own arguments, you have three options.
Pass it in as another argument, reference it as `pkg::fun()`, or move it into a
package.

**`package_function(package, symbol)`** names a function in an installed package
instead. It records a hash of that function's code. A worker process refuses to
run if the definition it loads does not match. Use it for production runs.

```{r package-function, message = FALSE}
fn <- package_function("batchit", "where_to_write_output")

names(fn)
fn$formal_names
```

The hash is deliberately narrow. It covers the target function's own body and
formals, and nothing else. A changed helper it calls, a changed constant, and a
different dependency version are all outside it. A match proves that the body and
the formals are the definition the parent R session hashed. It does not prove
that behaviour is identical. The hash ignores comments and whitespace, so an
installed package and a `devtools::load_all()` tree agree on identical code.

`symbol` may name an internal function. It does not have to be exported.

## Name every argument, every time

For every item, batchit checks that you named **every** formal `fn` has,
including ones with a default, and nothing else. It checks in the parent R
session before dispatching, and again in the worker process. Leaving out `scale`
here is an error, even though `scale` has a default:

```{r named-args, error = TRUE, message = FALSE}
run_and_collect(
  fn = function(x, scale = 1) x * scale,
  items = list(list(x = 2)),
  n_workers = 1
)
```

Name `scale` in the item and the same call succeeds:

```{r named-args-ok, message = FALSE}
run_and_collect(
  fn = function(x, scale = 1) x * scale,
  items = list(list(x = 2, scale = 10)),
  n_workers = 1
)
```

A named default is what makes a forgotten argument visible. An argument you meant
to set, but left out, otherwise looks exactly like one you left at its default
deliberately. For the same reason, `fn` MUST NOT take `...`.

## What the atomic-write guarantee covers

batchit stages every one of an item's declared outputs before it replaces any
final file. It then replaces the final files one at a time, each by a rename
within the destination directory. Last, it writes a marker recording the commit.
The guarantee is different in each of those three phases.

**Before the first replacement: fully covered.** The item can fail here in four
ways:

- an error in `fn`;
- a `"return"` value whose names do not match the declared outputs;
- a declared output a `"staged_writer"` never wrote;
- a failure serializing one of the staged files.

After any of those four, **no declared output** is touched. Whatever was at those
paths stays exactly as it was, and if nothing was there, nothing appears. batchit
removes the item's own previous marker before the first replacement, so that
marker does not survive even here.

**During the replacement loop: not transactional.** The worker process can be
killed partway through the loop, or a rename can fail. Some of the item's outputs
are then new while others are still old. batchit never rolls back a file it
already renamed. An individual file is not left half-written, since rename
replacement is atomic under the filesystem semantics batchit supports, but the
*set* can be torn.

**After the marker is written: committed, but the call may still report failure.**
The worker serializes its result envelope only after it commits. That
serialization can fail, or the worker can be killed in that window. Every final
file and the marker are then correctly in place, while the call reports the item
as failed.

**A timeout is not confined to one phase.** `timeout` measures the worker
process's whole lifetime, so it can land in any of the three phases. batchit's own
source notes that a timeout can kill a worker mid-commit. Do not read "timeout"
as a pre-replacement failure.

The marker is what tells a complete commit from a torn one. It is a small file
named `.batchit__<item id>`, written only after every replacement has succeeded.
A torn set has no valid marker vouching for it. batchit places one per item, in
the directory of the item's lexicographically first declared output path. That is
beside the outputs when they share a directory, and in one of them when they do
not. Do not create, read or depend on that file yourself; batchit manages it.
Because the name is built from the item id, an id used with either file-writing
function MUST NOT contain `/` or `\`.

**Not a skip mechanism.** batchit never treats an existing output file as a
reason to skip an item. It schedules every item, subject only to validation or an
earlier failure stopping the call first.

**Per item, not per call.** When one item fails the call stops, but items that
already committed keep their files. batchit does not undo them.

## When something fails

If an item's worker process raises an error, exits unexpectedly, or runs past
`timeout`, the call stops with an R error. It does not continue past the failure.
It never returns a partial result with an error object in the failed slot. Work
already running may finish before the call unwinds.

```{r failure, error = TRUE}
run_and_collect(
  fn = function(x) stop("this item cannot run"),
  items = list(fit_03 = list(x = 1)),
  n_workers = 1
)
```

Item ids come from the names of `items`, or from `ids` when streaming. They label
progress and error messages. Name your items, so that a message reads
`item 'fit_03' exceeded the timeout` rather than `item '3'`.

`run()`, `run_and_collect()` and `run_and_write_files_atomically()` print the
failing worker process's combined stdout and stderr first, so the cause is
visible. That log tail is the last 64,000 bytes, and at most 100 lines.
`stream_from_parent_and_write_files_atomically()` reports the error that came back
through `mirai`. It has no equivalent captured-log tail.

On success, a worker process's stdout and stderr are discarded. Warnings are the
exception. batchit carries back up to the first 100 warnings per item. It
re-raises them in the parent R session, labelled with the item id:

```{r warnings}
run_and_collect(
  fn = function(x) {
    warning("value looks unusual")
    x
  },
  items = list(fit_01 = list(x = 1)),
  n_workers = 1
)
```

**Only a completed item's warnings reach you.** If an item warns and then fails,
the error envelope carries no warnings, and the call raises the error alone. Those
warnings stay in the worker's captured output. The three non-streaming functions
print that output when they report the failure. Read that log; do not wait for an
R warning.

`timeout` is 6 hours by default, per item. Pass `Inf` to disable it.

## Two settings worth knowing

**`dev_path`** loads a package source tree in each worker process with
`devtools::load_all()`, instead of using the installed package. Which tree it must
name depends on `fn`:

- with a `package_function()` target, it is that package's source tree. Load the
  same tree in the parent R session before you build the descriptor. The hash the
  parent records then matches the code the worker process loads.
- with an inline function, there is no consumer package to load, and `dev_path`
  can only name batchit's own source tree. It is useful when you develop batchit
  itself.

Leave it `NULL` to use installed packages. A path that does not exist, or that
does not hold the expected package, is an error. It never falls back to the
installed version quietly.

**Thread counts.** batchit does not set BLAS or `data.table` thread counts for
you. If `fn` is itself multi-threaded, lower its thread count when you run many
worker processes. Otherwise they compete for the same cores.

## Writing one file atomically, with no dispatch

`write_qs2_atomically(object, path, ...)` is the commit engine's
rename-into-place step, on its own. It writes to a uniquely-named temporary file
in the destination's own directory, then renames it over the destination. No
workers, no items, no `fn`. It returns the destination path, invisibly.

```{r atomic}
cars_path <- write_qs2_atomically(mtcars, file.path(tempdir(), "cars.qs2"))

basename(cars_path)
nrow(qs2::qs_read(cars_path))
```

Use it wherever an interrupted write would leave a truncated file. A later read
halts on that file, and a long pipeline then restarts from the beginning. The
destination is left either absent or complete, never partial.

It is independent of `run_and_write_files_atomically()`. It writes no marker. Its
temporary file carries no dispatch attempt token, and that function's failure
cleanup does not sweep it. Read the three things it does **not** promise in
`?write_qs2_atomically` before you rely on it:

- it is not durability, because a rename is not an `fsync`, so a power loss can
  still lose the file;
- it is not a lock, because two concurrent writers each produce a complete file
  and the last rename wins;
- a `SIGKILL` leaves the temporary file behind, since `on.exit()` cannot run.

## Where to go next

- `?run`, `?run_and_collect`, `?run_and_write_files_atomically` and
  `?stream_from_parent_and_write_files_atomically` document every argument.
- `?package_function` and `?where_to_write_output` document the two helpers.
- `?write_qs2_atomically` documents the standalone atomic writer.
- The [reference index](https://papadopoulos-lab.github.io/batchit/reference/index.html)
  lists every exported function on one page.
